---
title: "Data Viz Final: 2023 MLB Ohtani Pitching"
subtitle: "By Alan Guo"
format: 
  revealjs: 
    incremental: true  
    slide-number: true
editor: visual
---

## Background {auto-animate="true"}

-   Sports is an industry where statistics are abundant, and data analysis is vital.

-   While every sport has its fair share of data collection, baseball stands out as a data-based sport

-   Sabermetrics - the empirical analysis of baseball, especially baseball statistics that measure in-game activity.

-   Shohei Ohtani is a Japanese baseball player in the league that is widely considered to best player right now, and the only two-way player (bats and pitches.)

-   I will analyze his pitching stats for the 2023 MLB season.

## The Data  {auto-animate="true"}

I am using data from a website called baseballsavant. They have literally every data from every at bat from every game. Some stats include type of pitch, velocity, pitch placement, result of at-bat, etc. It is a truly comprehensive data-set with virtually no missing data.

-   <https://baseballsavant.mlb.com/statcast_search> (here is search page)

-   <https://tinyurl.com/ohtanipitch> (here is the exact search, and click on download data as csv on top right of table)

## Goal  {auto-animate="true"}

-   Although analyzing Ohtani pitching stats, side goal is to just analyze pitching trends in the MLB.

-   Data very immense if I include all pitchers

-   Although this means sample size is not representative of MLB, it offers general pitching trends that are true for all pitchers

-   Graphic Set 1 focuses on Ohtani's pitching arsenal

-   Graphic Set 2 focuses on the strikezone placement

-   Graphic Set 3 focuses on pitches based on counts

## Some context for understanding baseball  {auto-animate="true"}

-   Pitches are classified as fastballs or breaking balls (balls that curve and deceive)

-   Fastballs: A 4-seamer is your classic fastball that goes straight, a cutter is a fastball that cuts horizontally toward the pitcher's glove side, a sinker cuts the other way vs a cutter, and a splitter is a fastball that cuts down

-   Breaking Balls: Curveball is slow and curves down, a slider cuts horizontally toward the pitcher's glove side, and a sweeper is just a more extreme version of the slider

## Graphics Set 1: Pitch Type Distribution {auto-animate="true"}

```{r}
library(tidyverse)
library(baseballr)
library(RColorBrewer)

ohtani <- read_csv("data/ohtani.csv", show_col_types = FALSE)
```

```{r}
#| echo: false
ggplot(ohtani, aes(x = pitch_type)) +
  geom_bar(aes(fill = type)) +
  geom_text(stat = "count", 
            aes(label = after_stat(count)),
            vjust = -0.5, hjust = 0.5) +
  scale_x_discrete(labels = c("Curveball",
                                "Cutter*",
                                "4-Seam*",
                                "Splitter",
                                "Sinker",
                                "Slider",
                                "Sweeper*")) +
  scale_fill_brewer(name = "Result", 
                    labels = c("Ball",
                               "Strike",
                               "In Play"), 
                    palette = "Set3") +
  xlab("Pitch Type") +
  ylab(NULL) +
  labs(title = "Pitch counts",
       caption = "* indicates primary pitch") + 
  scale_y_continuous(limits = c(0,900)) +
  theme_minimal() + 
  theme(legend.text = element_text(size = 6),
        axis.text.x = element_text(size = 6),
        axis.title.x = element_text(size = 10),
        plot.caption = element_text(color = "gray40"))
  
```

## Graphics Set 1: Pitch Type Distribution {auto-animate="true"}

```{r}
#| echo: false
ggplot(ohtani, aes(x = pitch_type)) +
  geom_bar(aes(fill = type), position = "fill") +
  scale_x_discrete(labels = c("Curveball",
                                "Cutter*",
                                "4-Seam*",
                                "Splitter",
                                "Sinker",
                                "Slider",
                                "Sweeper*")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_brewer(name = "Result", 
                    labels = c("Ball",
                               "Strike",
                               "In Play"), 
                    palette = "Set3") +
  xlab("Pitch Type") +
  ylab(NULL) +
  labs(title = "Pitch Result Percentages",
       caption = "* indicates primary pitch") + 
  theme_minimal() +
  theme(legend.text = element_text(size = 6),
        axis.text.x = element_text(size = 6),
        axis.title.x = element_text(size = 10),
        plot.caption = element_text(color = "gray40"))
```

## Graphics Set 1: Pitch Type Distribution {auto-animate="true"}

```{r}
#| echo: false
ggplot(ohtani, aes(x = pfx_x, y = pfx_z)) +
  scale_color_brewer(name = "Pitch Type",
                     labels = c("Curveball",
                                "Cutter",
                                "4-Seam",
                                "Splitter",
                                "Sinker",
                                "Slider",
                                "Sweeper"),
                     palette = "Set3") + 
  geom_point(aes(color = pitch_type)) + 
  xlab("Horizontal Break (ft)") +
  ylab("Verticle Break (ft)") +
  theme_minimal() +
  labs(title = "Pitch Movement by Pitch Type") +
  theme()
```

## Takeaway {.smaller auto-animate="true"}

-   Primary Pitches: 4-Seam Fastball, Cutter, Sweeper

-   Secondary Pitches: Splitter, Sinker, Curveball

-   Primary pitches tend to have a higher strike rate. Makes sense because those are more practiced and used

-   Sweeper breaks right the most, cutter a bit to right, curveball down. Sinker most to left

-   Surprisingly, his 4-seam has verticle break instead of expected middle, and his splitter has left break instead of down

-   Sweeper and Cutter has most variation in breaks, and also the most in-play percentage. More unpredictable = less control = ball hit

## Graphics Set 2: Strike Zone {auto-animate="true"}

```{r}
#| echo: false
ggplot(ohtani, aes(x = plate_x, y = plate_z)) +
  geom_point(alpha = 0.4, aes(color = type)) +
  geom_density_2d(bins = 10, color = "grey40", alpha = 0.7) +
  geom_rect(mapping = aes(ymax = 3.56, ymin = 1.6, 
                          xmax = -1, xmin = 1), 
            alpha = 0, 
            size=1.2,
            colour = "black") +
  scale_x_continuous(limits = c(-3,3)) +
  scale_y_continuous(limits = c(0,4)) +
  scale_color_brewer(name = "Result", 
                    labels = c("Ball",
                               "Strike",
                               "In Play"), 
                    palette = "Set1") +
  xlab("Plate Horizontal Location (ft)") +
  ylab("Plate Vertical Location (ft)") +
  labs(title = "Pitch Distribution") +
  theme_minimal()
```

## Graphics Set 2: Strike Zone {auto-animate="true"}

```{r}
#| echo: false
hit <- c("single", "double", "home_run")

ohtani_hit <- ohtani %>%
  filter(events %in% hit)

ggplot(ohtani_hit, aes(x = plate_x, y = plate_z, color = pitch_type)) +
  geom_point() + 
  geom_rect(mapping = aes(ymax = 3.56, ymin = 1.6, 
                          xmax = -1, xmin = 1), 
            alpha = 0, 
            size=1.2,
            colour = "black") +
  scale_color_brewer(name = "Pitch Type",
                       labels = c("Curveball",
                                "Cutter",
                                "4-Seam",
                                "Splitter",
                                "Sinker",
                                "Sweeper"),
                     palette = "Set3") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(1,4)) +
  xlab("Plate Horizontal Location (ft)") +
  ylab("Plate Vertical Location (ft)") +
  labs(title = "Hits with Pitch Type") +
  theme_minimal()
```

## Takeaway {.smaller auto-animate="true"}

-   The distribution of the pitches go diagonally from top-left to bottom right, implying a right-handed pitcher (arm motion goes top-right to bottom left, inverse of catcher's view)

-   Throws outside of the box are mostly outside and low, sign of sweepers, cutters, curveballs

-   Naturally most strikes are in the strikezone, most balls are outside

-   The majority of hits off of his pitches are in the middle or lower strikezone, which makes sense since hitting a lower ball usually means higher angled hits which can get lift off of the ground.

-   No clear trend between pitch type and hits, although some are visually off-position (lots of sweepers going middle when they should be bottom right)

## Graphics Set 3: Count Analysis  {auto-animate="true"}

```{r}
ohtani_count <- ohtani %>%
  mutate(count = paste(balls, strikes, sep = "-"))

ggplot(ohtani_count, aes(x = "")) +
  geom_bar(aes(fill = pitch_name), position = "fill") +
  coord_polar("y", start = 0) +
  facet_wrap(~count) +
  scale_fill_brewer(name = "Pitch Type",
                    palette = "Set3") +  # Change the palette if needed
  labs(title = "Pitch Types by Count") +
  theme_minimal() + 
  theme(legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.2, "cm"),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank())

```

## Graphics Set 3: Count Analysis {auto-animate="true"}

```{r}
ggplot(ohtani_count, aes(x = "")) +
  geom_bar(aes(fill = type), position = "fill") +
  coord_polar("y", start = 0) +
  facet_wrap(~count) +
  scale_fill_brewer(name = "Result",
                    labels = c("Ball",
                               "Strike",
                               "In Play"),
                    palette = "Set3") +  # Change the palette if needed
  labs(title = "Pitch Results by Count") +
  theme_minimal() + 
  theme(legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.2, "cm"),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank())
```

## Takeaway  {auto-animate="true"}

-   The more behind on the count (more balls than strikes) the more fastballs he pitches

-   The more ahead on the count (more strikes than balls) or neutral, he throws with a lot more variation

-   He only utilizes his splitter in high quantity when he's ahead with 2 strikes

-   Naturally he throws more strikes when he's behind, and more balls when he's ahead

## Conclusion  {auto-animate="true"}

-   Ohtani carries a wide arsenal of pitches but primarily uses 4-seamers, cutters, and sweepers

-   As a right handed pitcher, he tends to distribute his pitches diagonally respective to his arm motion (top left to bottom right in catcher POV)

-   He pitches a lot more safely when he's behind on the count, and with a lot more variation when he's ahead

-   This data viz doesn't compare him to other pitchers so there's not really a benchmark to compare him to, but this visualization is a great summary of Ohtani's pitching trends
